---
title: "FusioMR Software Tutorial"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{FusioMR Tutorial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "##",
  prompt = FALSE,
  warning = FALSE
)
```

```{r load, include = FALSE}
library(FusioMR)
```

This vignette demonstrates how to use `FusioMR` for single- and
multi-outcome Mendelian randomization with GWAS summary statistics.
For installation instructions, package overview, and a description of
input / output formats, see the
[README](https://github.com/kangbw702/FusioMR).

The example datasets used below are available in the
[examples/data/](https://github.com/kangbw702/FusioMR/tree/main/examples/data)
folder of the GitHub repository.

# FusioMR_s: Single-outcome MR model allowing pleiotropy

## Single Exposure, Single Outcome: `seso_uhp_only`

The `seso_uhp_only` model handles the most common MR setup: one exposure,
one outcome, and uncorrelated horizontal pleiotropy as the main concern. The
following is the function usage example with a pre-generated dataset
simulated under the limited-IV TWMR setting (exposure GWAS sample size = 300,
outcome GWAS sample size = 20,000, true causal effect $\beta = 0$).

```{r seso-uhp-load}
d1 <- readRDS("data/seso_uhp_only_example.rds")
fit1 <- fusiomr(d1$b_exp,
                d1$se_exp,
                d1$b_out,
                d1$se_out,
                model = "seso_uhp_only")

cat(sprintf("The causal effect estimate is %.4f\n", fit1$est))
cat(sprintf("The standard error is %.4f\n", fit1$se))
cat(sprintf("The two-sided p-value is %.4g\n", fit1$pval))
cat(sprintf("The 95%% CI is [%.4f, %.4f]\n", fit1$ci[1], fit1$ci[2]))
```

## Single exposure, Single outcome with CHP: `seso_with_chp`

When some IVs are suspected of having correlated horizontal pleiotropy, use
`seso_with_chp`. We use a complex-trait example for model illustration
(exposure and outcome GWAS sample sizes both 20,000, true causal effect
$\beta = 0$).

```{r seso-chp-fit}
d2 <- readRDS("data/seso_with_chp_example.rds")
fit2 <- fusiomr(d2$b_exp,
                d2$se_exp,
                d2$b_out,
                d2$se_out,
                model = "seso_with_chp")

cat(sprintf("The causal effect estimate is %.4f\n", fit2$est))
cat(sprintf("The standard error is %.4f\n", fit2$se))
cat(sprintf("The two-sided p-value is %.4g\n", fit2$pval))
cat(sprintf("The 95%% CI is [%.4f, %.4f]\n", fit2$ci[1], fit2$ci[2]))
```

# FusioMR_m: Multi-outcome MR models accounting for shared pleiotropy

## Single exposure, Two outcomes: `semo`

`semo` jointly analyzes one exposure and two outcomes. The input format
differs from the single-outcome models: `b_out` and `se_out` must be
`K x 2` matrices, where column $j$ contains summary statistics for
outcome $j$. We use a TWMR-style example with one shared exposure and two
outcomes for illustration (exposure sample size = 300, outcome 1 sample
size = 80,000, outcome 2 sample size = 20,000, true causal effects
$\beta_1 = \beta_2 = 0.2$).

```{r semo-fit}
d3 <- readRDS("data/semo_example.rds")
fit3 <- fusiomr(d3$b_exp,
                d3$se_exp,
                d3$b_out,
                d3$se_out,
                model = "semo")

cat("Outcome 1:\n")
cat(sprintf("The causal effect estimate is %.4f\n", fit3$est[1]))
cat(sprintf("The standard error is %.4f\n", fit3$se[1]))
cat(sprintf("The two-sided p-value is %.4g\n", fit3$pval[1]))
cat(sprintf("The 95%% CI is [%.4f, %.4f]\n", fit3$ci[1, 1], fit3$ci[1, 2]))

cat("Outcome 2:\n")
cat(sprintf("The causal effect estimate is %.4f\n", fit3$est[2]))
cat(sprintf("The standard error is %.4f\n", fit3$se[2]))
cat(sprintf("The two-sided p-value is %.4g\n", fit3$pval[2]))
cat(sprintf("The 95%% CI is [%.4f, %.4f]\n", fit3$ci[2, 1], fit3$ci[2, 2]))
```

## Two exposures, Two outcomes: `memo`

`memo` jointly analyzes two correlated exposures and two correlated
outcomes, accounting for both UHP and CHP. All four summary-statistics
inputs (`b_exp`, `se_exp`, `b_out`, `se_out`) must be `K x 2` matrices,
where column $j$ corresponds to exposure $j$ (or outcome $j$). We use a
two-exposure, two-outcome example for illustration (exposure sample size = 300, 
outcome 1 sample size = 50,000, outcome 2 sample size = 50,000, true causal effects
$\beta_1 = \beta_2 = 0.2$).

```{r memo-fit}
d4 <- readRDS("data/memo_example.rds")
fit4 <- fusiomr(d4$b_exp,
                d4$se_exp,
                d4$b_out,
                d4$se_out,
                model = "memo")

cat("Outcome 1:\n")
cat(sprintf("The causal effect estimate is %.4f\n", fit4$est[1]))
cat(sprintf("The standard error is %.4f\n", fit4$se[1]))
cat(sprintf("The two-sided p-value is %.4g\n", fit4$pval[1]))
cat(sprintf("The 95%% CI is [%.4f, %.4f]\n", fit4$ci[1, 1], fit4$ci[1, 2]))

cat("Outcome 2:\n")
cat(sprintf("The causal effect estimate is %.4f\n", fit4$est[2]))
cat(sprintf("The standard error is %.4f\n", fit4$se[2]))
cat(sprintf("The two-sided p-value is %.4g\n", fit4$pval[2]))
cat(sprintf("The 95%% CI is [%.4f, %.4f]\n", fit4$ci[2, 1], fit4$ci[2, 2]))
```

# Advanced Parameter Setting: `parameter_control()`

For most users, the defaults are sufficient and `parameter_control()`
does not need to be called explicitly. When advanced parameter setting is needed, pass a
customized control list to `fusiomr()`:

A summary of the available tuning parameters:

| Argument             | Purpose |
|----------------------|--------------------------------------------------------------------------------------------|
| `niter`              | Number of Gibbs iterations (default 20000).                                                |
| `burnin_prop`        | Burn-in proportion in [0, 1) (default 0.5).                                                |
| `c_gamma`, `c_theta` | Prior weight per IV for $\sigma^2_\gamma$ and $\sigma^2_\theta$. Larger values make the prior more concentrated around its mean. |
| `kappa_gamma`, `kappa_theta` | Conservative inflation multipliers (>= 1) for the prior means.                     |
| `Kmin`, `Kmax`       | Floor and cap on $K$ when computing prior shape parameters.                                |
| `rho_ov`             | Sampling correlation between exposure and outcome due to sample overlap, in [-1, 1].       |
| `z_thresh`           | Optional $|Z_\gamma|$ selection threshold (winner's-curse correction). For p = 5e-8 two-sided, use `qnorm(1 - 5e-8/2)`. |
| `trim`               | Tail probability for winsorized moments (default 0.1).                                     |
| `hybrid`             | Logical; enable hybrid empirical-Bayes priors (see next section).                          |
| `kappa_hybrid`       | Pooling strength for hybrid mode; larger values shrink the local moment toward the global. |
| `global_mean_gamma`, `global_mean_theta` | Global EB centers (scalars), required for hybrid mode in `seso_*` models. |
| `global_Sigma_gamma`, `global_Sigma_theta` | Global EB centers (2x2 matrices), required for hybrid mode in `semo` / `memo`. |

See `?parameter_control` for full details.


```{r control-example, eval = FALSE}
ctrl <- parameter_control(
  niter = 30000,                   # longer MCMC
  z_thresh = qnorm(1 - 5e-8 / 2),  # winner's-curse correction
  rho_ov  = 0.2                    # sample overlap correction
)

fit <- fusiomr(b_exp, 
               se_exp, 
               b_out, 
               se_out,
               model = "seso_uhp_only", 
               control = ctrl)
```



## Hybrid empirical-Bayes priors (`hybrid = TRUE`)

By default, FusioMR sets the variance priors using a method-of-moments
estimate computed from the input IVs alone (the *local* moment).
In gene-region-specific TWMR analyses where the number of IVs per region
is small, the local moment can be noisy. The hybrid mode pools the local
moment with a global empirical-Bayes center estimated from a larger
reference set.

Setting `hybrid = TRUE` requires the user to supply the appropriate
global center(s) for the chosen model:

| Model           | Required global argument(s)                           | Type            |
|-----------------|-------------------------------------------------------|-----------------|
| `seso_uhp_only` | `global_mean_gamma`, `global_mean_theta`              | scalar, scalar  |
| `seso_with_chp` | `global_mean_gamma`, `global_mean_theta`              | scalar, scalar  |
| `semo`          | `global_mean_gamma`, `global_Sigma_theta`             | scalar, 2x2 mat |
| `memo`          | `global_Sigma_gamma`, `global_Sigma_theta`            | 2x2 mat, 2x2 mat |

The typical workflow is two-pass:

1. **Estimate the global centers** by running FusioMR on a large
   reference set of exposures (e.g., all genes in a TWMR scan) and
   summarizing the per-exposure local moments (e.g., taking the median
   across exposures).
2. **Refit the target exposure** with `hybrid = TRUE`, passing the
   global centers to `parameter_control()`.

### pseudo-code with hybrid priors

```{r hybrid-seso, eval = FALSE}
# After computing global_mean_gamma and global_mean_theta from a reference scan:

# seso example
ctrl_hybrid <- parameter_control(
  hybrid            = TRUE,
  kappa_hybrid      = 5,
  global_mean_gamma = global_mean_gamma,   # your global estimate (scalar)
  global_mean_theta = global_mean_theta    # your global estimate (scalar)
)

fit_hybrid <- fusiomr(b_exp, 
                      se_exp, 
                      b_out, 
                      se_out,
                      model = "seso_uhp_only", 
                      control = ctrl_hybrid)
```

If `hybrid = TRUE` is set but the required global argument(s) are
missing or have the wrong shape, `fusiomr()` stops with an informative
error message before running the sampler.

# Getting help

```{r help, eval = FALSE}
?fusiomr
?parameter_control
```